```{r}

```

---
title: "Substance Abuse"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("ggplot2")
install.packages("readr")



```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
library(rmarkdown)
library(ggplot2)
library(readr)
```

# Substance Abuse

Le Quan Thein - 2848350
Imke de Jonge - 2858894

Tutorial Group 6

Jack Fitzgerald

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Drug abuse continues to pose a significant public health and societal challenge in urban centers across Europe. Among the various methods used to monitor and understand patterns of drug consumption, wastewater-based epidemiology has emerged as a scientifically robust, objective, and non-invasive approach. By analyzing drug residues in municipal wastewater, researchers can estimate real-time community-level drug usage, offering valuable insights that are often difficult to capture through surveys or law enforcement data alone.

This report focuses specifically on MDMA (commonly known as ecstasy)—a powerful psychoactive stimulant associated with recreational use, particularly in nightlife settings. Understanding MDMA consumption patterns is crucial, not only because of its addictive potential and health risks, but also due to its role in shaping broader drug policy and public health responses. Using data from three major Dutch cities—Utrecht, Amsterdam, and Eindhoven—this analysis aims to identify spatial and temporal trends in MDMA usage, assess the impact on local populations, and inform evidence-based interventions. The relevance of this topic lies in its contribution to targeted harm reduction strategies and more effective urban drug policy planning.

# Part 2 - Data Sourcing

## 2.1 Load in the data

```{r loading_data}
dataset <- scoremdma20112017
drugs_NL <- scoremdma20112017[108:112,]
df_long <- drugs_NL
df_plot_ams <- df_long %>% filter(city == "Amsterdam")
df_plot_utr <- df_long %>% filter(city == "Utrecht")
df_plot_eid <- df_long %>% filter(city == "Eindhoven")

```

## 2.2 Provide a short summary of the dataset(s)

```{r}
head(scoremdma20112017)
```

The dataset comprises measurements of five illicit drug residues in wastewater, collected during a one-week monitoring campaign each year from **2011 to 2017**. It spans **143 wastewater treatment plants** across **120 cities in 37 countries**, with each site sampled at least once during the period.

Published by the Swiss Federal Institute of Aquatic Science and Technology (**Eawag**) in collaboration with SCORE, the dataset includes per-capita loads of:

-   Amphetamine

-   Benzoylecgonine (a cocaine metabolite)

-   MDMA

-   Methamphetamine

-   THC-COOH (a cannabis metabolite)

It provides population-normalized mass loads (mg/day per 1,000 inhabitants) of these substances in raw municipal wastewater during annual sampling weeks.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

### **Variable Descriptions:**

-   **`id`**: A unique identifier for each data entry corresponding to a specific city and year.

-   **`city`**: The name of the city where the wastewater sample was collected (e.g., Amsterdam, Utrecht, Eindhoven).

-   **`country`**: The country code representing where the city is located (e.g., NL for the Netherlands).

-   **`year`**: The calendar year during which the wastewater sample was analyzed.

-   **`population`**: The estimated number of people served by the wastewater treatment plant in the respective city

-   **`mean`**: This represents the average daily amount of MDMA detected in municipal wastewater, standardized per 1,000 inhabitants. It is expressed in milligrams per day per 1,000 people (mg/day/1000 people) and reflects community-level MDMA consumption during the sampling period.

<!-- -->

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}
dataset <- scoremdma20112017
drugs_NL <- scoremdma20112017[108:112,]
df_long <- drugs_NL
df_plot_ams <- df_long %>% filter(city == "Amsterdam")
df_plot_utr <- df_long %>% filter(city == "Utrecht")
df_plot_eid <- df_long %>% filter(city == "Eindhoven")
```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

**Annual Percentage Change in MDMA Load**

```{r gen_var1}
df <- drugs_NL

df <- df %>%
       mutate(
             pct_2011_2012 = round(((`2012 mean` - `2011 mean`) / `2011 mean`) * 100, 1),
             pct_2012_2013 = round(((`2013 mean` - `2012 mean`) / `2012 mean`) * 100, 1),
             pct_2013_2014 = round(((`2014 mean` - `2013 mean`) / `2013 mean`) * 100, 1),
             pct_2014_2015 = round(((`2015 mean` - `2014 mean`) / `2014 mean`) * 100, 1),
             pct_2015_2016 = round(((`2016 mean` - `2015 mean`) / `2015 mean`) * 100, 1),
             pct_2016_2017 = round(
                   ifelse(
                         city == "Amsterdam",
                         ((`2017 mean` / `2015 mean`)^(1/2) - 1) * 100,  
                         ((`2017 mean` - ifelse(is.na(`2016 mean`), 1, `2016 mean`)) /
                             +                      ifelse(is.na(`2016 mean`), 1, `2016 mean`)) * 100
                     ),
                   1
               )
         )
library(ggplot2)


signed_log <- function(x) {
  sign(x) * log10(abs(x) + 1)
}


selected_cities <- c("Amsterdam", "Utrecht", "Eindhoven")


drugs_long_filtered <- drugs_long %>%
  filter(city %in% selected_cities)


ggplot(drugs_long_filtered, aes(x = year_start, y = signed_log(pct_change), color = city)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 2012:2017) +
  labs(
    title = "Yearly Percentage Change of MDMA Found in Sewage Water",
    x = "Year",
    y = "Signed log10 of % Change",
    color = "City"
  ) +
  theme_minimal()
```

**Population-Weighted Load Score**

```{r var_popweightedload}
df_plot_ams <- df_plot_ams %>%
  mutate(PopWeightedLoad = (mean/1000) * population)
print(df_plot_ams[,c("year", "mean", "population", "PopWeightedLoad")])
df_plot_ams <- df_plot_ams %>%
  filter(!is.na(mean))

df_plot_utr <- df_plot_utr %>%
  mutate(PopWeightedLoad = (mean/1000) * population)
print(df_plot_utr[,c("year", "mean", "population", "PopWeightedLoad")])

df_plot_eid <- df_plot_eid %>%
  mutate(PopWeightedLoad = (mean/1000) * population)
print(df_plot_eid[,c("year", "mean", "population", "PopWeightedLoad")])
```

## 3.3 Visualize temporal variation

```{r}
ggplot(data = df_plot_utr, aes(x = year, y = mean)) +
  geom_line(data = df_plot_utr, aes(x = year, y = mean, colour = "red", group=1)) +
  geom_point(data = df_plot_ams, aes(x = year, y = mean, colour = "blue",)) +
  geom_line(data = df_plot_eid, aes(x = year, y = mean, colour = "green", group=1)) +
  scale_color_identity(name="",
                       breaks=c("red","blue","green"
                                ),
                       labels=c("Utrecht", "Amsterdam","Eindhoven"),
                       guide="legend"
                       )+
  labs(y="mean", x="year") + 
  theme_minimal()

```

```{r}
#To get a close-up view on Utrecht's and Amsterdam's data
ggplot(data = df_plot_utr, aes(x = year, y = mean)) +
  geom_line(data = df_plot_utr, aes(x = year, y = mean, colour = "red", group=1)) +
  geom_point(data = df_plot_ams, aes(x = year, y = mean, colour = "blue",)) +
  geom_line(data = df_plot_eid, aes(x = year, y = mean, colour = "green", group=1)) +
  scale_color_identity(name="",
                       breaks=c("red","blue","green"
                                ),
                       labels=c("Utrecht", "Amsterdam","Eindhoven"),
                       guide="legend"
                       ) +
  coord_cartesian(ylim = c(0,900))+
  labs(y="mean", x="year") + 
  theme_minimal()
```

## 3.4 Visualize spatial variation

```{r visualise_map}
library(dplyr)
library(ggplot2)
library(sf)
library(rnaturalearth)
library(tidyverse)

score_mdma_2011_2017 <- readr::read_delim(
  "C:/Users/Thijmen/Downloads/score_mdma_wastewater_loads_2011_2017.csv",
  delim = ";",
  escape_double = FALSE,
  trim_ws = TRUE
)
drugs_NL <- score_mdma_2011_2017[108:112,]
df <- drugs_NL
colnames(df)[colnames(df) == "...1"] <- "city"
netherlands <- ne_countries(scale = "medium", country = "Netherlands", returnclass = "sf")
df$longitude  = c(4.9041, 5.4778, 5.0803, 4.8681, 5.1214) 
df$latitude  = c(52.3676, 51.4416, 52.0306, 52.0282, 52.0907)
df_sf <- st_as_sf(df, coords = c("longitude", "latitude"), crs = 4326)
ggplot() +
  geom_sf(data = netherlands, fill = "gray", color = "black") +
  geom_sf(data = df_sf, aes(color = `2017 mean`), size = 5) +
  scale_color_gradientn(
    colors = c("green", "orange", "red"),
    name = "MDMA (mg/ 1000 people/day)"
  ) +
  geom_text(
    data = df,
    aes(x = longitude, y = latitude, label = city),
    vjust = ifelse(df$city == "Oudewater", 1.5, -1.5),
  ) +
  coord_sf(xlim = c(4, 7), ylim = c(51, 53)) +
  labs(title = "Mean levels of MDMA in wastewater in 2017 (mg/1000 people/day") +
  theme_minimal()
```
The spatial visualization gives us an idea of where in the Netherlands MDMA is most commonly found in sewage water in the year 2017, the most recent year of the study. In the plot, we can see that especially Amsterdam and Eindhoven have a high concentration of MDMA in their sewage. This is relevant for the broader problem of drug abuse, because it explains where this drug is most commonly consumed, and where it is produced. We notice, for example, that Amsterdam has a much higher concentration of MDMA than Utrecht. This can mean that Utrecht has had a succesful prevention/intervention policy that could be implemented in Amsterdam as well. It also indicates that time, money and effort are better allocated to Amsterdam since the problem of MDMA use is higher there than it is in Utrecht. 
The spatial visualisation can be used as a starting point for further research, if data for more wastewater treatment plants will be available. Noord-Brabant, the province where Eindhoven is located, is known for having a large amount of MDMA manufacturing labs. If you compare the MDMA level found in Eindhoven to other wastewater treatment plants in Noord-Brabant, you can get a better idea of which places in the province have a higher density of drug manufacturing labs. 



## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
library(ggplot2)
data <- read.csv("cannabis_usage_by_group.csv")


ggplot(data, aes(x = Group, y = Cannabis.Usage....)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Cannabis Use in the Past Year by Age Group",
       x = "",
       y = "Percentage (%)") +
  theme_minimal()
```

The boxplot comparing cannabis use between two age groups—Sub Population (15–24) and Main Population (25+)—reveals clear differences in usage patterns. Young adults (15–24) exhibit a higher median rate of cannabis use over the past year, alongside greater variability. This suggests not only more frequent usage in this group but also a wider range of behaviors, with some subgroups showing especially elevated rates. The interquartile range for the 15–24 group is significantly broader, indicating that cannabis consumption among young people is more heterogeneous. In contrast, the 25+ group demonstrates a lower and more concentrated distribution of cannabis usage, implying more uniform and restrained behavior within this demographic. One possible explanation lies in social and developmental factors—young adults are more likely to experiment, be influenced by peer pressure, or engage in risk-taking behavior. The data may also reflect generational shifts in attitudes toward cannabis, with younger cohorts being more accepting or curious. From a public health or policy perspective, these findings underline the importance of targeted interventions aimed at youth, especially in preventive education or harm reduction strategies. Overall, age remains a strong determinant of drug-use patterns, and understanding these differences is essential for effective, age-sensitive health communication and policymaking.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings
The analysis of MDMA use based on wastewater data in three Dutch cities—Amsterdam, Utrecht, and Eindhoven—provides insights not only into local differences but also contributes to the broader societal debate on substance use. In this section, we reflect on our findings, link them to existing literature, and discuss the societal implications and limitations of our approach.

### Substance Abuse as a Broader Social Problem

As outlined in our problem motivation, substance abuse has far-reaching consequences: from increased healthcare costs and drug-related crime to public nuisance and mental health challenges. Data from CBS (2020, 2023) and RIVM (2024) confirm rising trends in the use of substances such as ecstasy and cannabis, especially among young adults. Our findings partially reflect this: Amsterdam—an international nightlife hub with a young, tourist-heavy demographic—shows a relatively stable or increasing trend in MDMA residues, suggesting that increased use among youth and visitors is indeed observable at the urban infrastructure level.

### Urban Differences and Policy Effects

The observed differences between the cities highlight the influence of social, demographic, and policy-driven factors. Utrecht shows an early decline following the introduction of the “weed pass” in 2011–2012, likely reflecting reduced access and social acceptance of drug use. Eindhoven displays a dramatic spike in 2014, which can be attributed to illegal drug-lab waste dumping rather than increased local use—an indicator of the production side of the drug problem. Amsterdam’s more permissive stance toward soft drugs, combined with high tourist inflow, may explain the city's consistent levels of MDMA detected in wastewater.

### Strengthening the Analysis with New Variables

We introduced two new variables: **Annual Percentage Change** and **Population-Weighted Load**. These provide greater analytical depth. The former reveals year-over-year changes, capturing volatility or stability in trends, while the latter allows for comparisons in absolute usage volume between cities of different sizes. These additions support more accurate interpretation and improve the ability to prioritize interventions based not only on per capita use but also on total urban consumption.

### Limitations and Recommendations

While wastewater analysis offers objective, anonymized, and large-scale data, it comes with limitations. It measures residues, not individual behaviors; it can be distorted by external events like dumping; and it lacks demographic detail about users. To address these gaps, future research could combine wastewater data with survey data or healthcare records. Additionally, Amsterdam's missing data for 2016 presents a gap in the time series that should be acknowledged when interpreting trends.

### Societal Implications

Our findings can help policymakers target prevention and intervention efforts more effectively—whether through enhanced monitoring in tourist-heavy areas, youth education campaigns, or stricter control of industrial waste associated with drug production. The analysis offers a clearer picture of **where** substance use is concentrated and **how** it evolves, which is crucial for designing tailored and evidence-based drug policies.

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list
Castagna, G. (2025, 24 maart). Cijfers drugs. Trimbos-instituut. https://www.trimbos.nl/kennis/cijfers/drugs/
CBS Statline. (z.d.). https://opendata.cbs.nl/statline/#/CBS/nl/dataset/85457NED/table?ts=1749653240915
Drugsgebruik | Volwassenen. (z.d.). Volksgezondheid en Zorg. https://www.vzinfo.nl/drugsgebruik/volwassenen?utm_source=
Hupkens, E. D. C. (2023, 13 december). Bijlagetabellen. Centraal Bureau Voor de Statistiek. https://www.cbs.nl/nl-nl/longread/statistische-trends/2023/drugsgebruik-en-mentale-gezondheid/bijlagetabellen
Illicit drugs in wastewater - SCORE initiative - SCORE_MDMA_wastewater_loads_2011_2017.csv - ERIC. (z.d.). https://opendata.eawag.ch/dataset/illicit-drugs-in-wastewater-score-initiative/resource/df0ea8b4-6720-476f-8de3-cdb9d5968ed6
Ministerie van Justitie en Veiligheid. (2024, 31 januari). Drugs. Government.nl. http://government.nl/topics/drugs#:~:text=In%20the%20Netherlands%20as%20elsewhere,who%20are%20addicted%20to%20drugs.
Netherlands, S. (2020, 4 september). One-quarter of population notice drug-related issues in own neighbourhood. Statistics Netherlands. https://www.cbs.nl/en-gb/news/2020/36/one-quarter-of-population-notice-drug-related-issues-in-own-neighbourhood
Use APA referencing throughout your document.
