```{r}

```

---
title: "Substance Abuse"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("ggplot2")
install.packages("readr")



```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
library(rmarkdown)
library(ggplot2)
library(readr)
```

# Substance Abuse

Le Quan Thein - 2848350

Tutorial Group 6

Include your tutorial lecturer's name

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Drug abuse continues to pose a significant public health and societal challenge in urban centers across Europe. Among the various methods used to monitor and understand patterns of drug consumption, wastewater-based epidemiology has emerged as a scientifically robust, objective, and non-invasive approach. By analyzing drug residues in municipal wastewater, researchers can estimate real-time community-level drug usage, offering valuable insights that are often difficult to capture through surveys or law enforcement data alone.

This report focuses specifically on MDMA (commonly known as ecstasy)—a powerful psychoactive stimulant associated with recreational use, particularly in nightlife settings. Understanding MDMA consumption patterns is crucial, not only because of its addictive potential and health risks, but also due to its role in shaping broader drug policy and public health responses. Using data from three major Dutch cities—Utrecht, Amsterdam, and Eindhoven—this analysis aims to identify spatial and temporal trends in MDMA usage, assess the impact on local populations, and inform evidence-based interventions. The relevance of this topic lies in its contribution to targeted harm reduction strategies and more effective urban drug policy planning.

# Part 2 - Data Sourcing

## 2.1 Load in the data

```{r loading_data}
dataset <- scoremdma20112017
drugs_NL <- scoremdma20112017[108:112,]
df_long <- drugs_NL
df_plot_ams <- df_long %>% filter(city == "Amsterdam")
df_plot_utr <- df_long %>% filter(city == "Utrecht")
df_plot_eid <- df_long %>% filter(city == "Eindhoven")

```

## 2.2 Provide a short summary of the dataset(s)

```{r}
head(scoremdma20112017)
```

The dataset comprises measurements of five illicit drug residues in wastewater, collected during a one-week monitoring campaign each year from **2011 to 2017**. It spans **143 wastewater treatment plants** across **120 cities in 37 countries**, with each site sampled at least once during the period.

Published by the Swiss Federal Institute of Aquatic Science and Technology (**Eawag**) in collaboration with SCORE, the dataset includes per-capita loads of:

-   Amphetamine

-   Benzoylecgonine (a cocaine metabolite)

-   MDMA

-   Methamphetamine

-   THC-COOH (a cannabis metabolite)

It provides population-normalized mass loads (mg/day per 1,000 inhabitants) of these substances in raw municipal wastewater during annual sampling weeks.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

### **Variable Descriptions:**

-   **`id`**: A unique identifier for each data entry corresponding to a specific city and year.

-   **`city`**: The name of the city where the wastewater sample was collected (e.g., Amsterdam, Utrecht, Eindhoven).

-   **`country`**: The country code representing where the city is located (e.g., NL for the Netherlands).

-   **`year`**: The calendar year during which the wastewater sample was analyzed.

-   **`population`**: The estimated number of people served by the wastewater treatment plant in the respective city

-   **`mean`**: This represents the average daily amount of MDMA detected in municipal wastewater, standardized per 1,000 inhabitants. It is expressed in milligrams per day per 1,000 people (mg/day/1000 people) and reflects community-level MDMA consumption during the sampling period.

<!-- -->

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}
dataset <- scoremdma20112017
drugs_NL <- scoremdma20112017[108:112,]
df_long <- drugs_NL
df_plot_ams <- df_long %>% filter(city == "Amsterdam")
df_plot_utr <- df_long %>% filter(city == "Utrecht")
df_plot_eid <- df_long %>% filter(city == "Eindhoven")
```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

**Annual Percentage Change in MDMA Load**

```{r gen_var1}

```

**Population-Weighted Load Score**

```{r var_popweightedload}
df_plot_ams <- df_plot_ams %>%
  mutate(PopWeightedLoad = (mean/1000) * population)
print(df_plot_ams[,c("year", "mean", "population", "PopWeightedLoad")])
df_plot_ams <- df_plot_ams %>%
  filter(!is.na(mean))

df_plot_utr <- df_plot_utr %>%
  mutate(PopWeightedLoad = (mean/1000) * population)
print(df_plot_utr[,c("year", "mean", "population", "PopWeightedLoad")])

df_plot_eid <- df_plot_eid %>%
  mutate(PopWeightedLoad = (mean/1000) * population)
print(df_plot_eid[,c("year", "mean", "population", "PopWeightedLoad")])
```

## 3.3 Visualize temporal variation

```{r}
ggplot(data = df_plot_utr, aes(x = year, y = mean)) +
  geom_line(data = df_plot_utr, aes(x = year, y = mean, colour = "red", group=1)) +
  geom_point(data = df_plot_ams, aes(x = year, y = mean, colour = "blue",)) +
  geom_line(data = df_plot_eid, aes(x = year, y = mean, colour = "green", group=1)) +
  scale_color_identity(name="",
                       breaks=c("red","blue","green"
                                ),
                       labels=c("Utrecht", "Amsterdam","Eindhoven"),
                       guide="legend"
                       )+
  labs(y="mean", x="year") + 
  theme_minimal()

```

```{r}
#To get a close-up view on Utrecht's and Amsterdam's data
ggplot(data = df_plot_utr, aes(x = year, y = mean)) +
  geom_line(data = df_plot_utr, aes(x = year, y = mean, colour = "red", group=1)) +
  geom_point(data = df_plot_ams, aes(x = year, y = mean, colour = "blue",)) +
  geom_line(data = df_plot_eid, aes(x = year, y = mean, colour = "green", group=1)) +
  scale_color_identity(name="",
                       breaks=c("red","blue","green"
                                ),
                       labels=c("Utrecht", "Amsterdam","Eindhoven"),
                       guide="legend"
                       ) +
  coord_cartesian(ylim = c(0,900))+
  labs(y="mean", x="year") + 
  theme_minimal()
```

## 3.4 Visualize spatial variation

```{r visualise_map}
library(dplyr)
library(ggplot2)
library(sf)
library(rnaturalearth)
library(tidyverse)

score_mdma_2011_2017 <- readr::read_delim(
  "C:/Users/Thijmen/Downloads/score_mdma_wastewater_loads_2011_2017.csv",
  delim = ";",
  escape_double = FALSE,
  trim_ws = TRUE
)
drugs_NL <- score_mdma_2011_2017[108:112,]
df <- drugs_NL
colnames(df)[colnames(df) == "...1"] <- "city"
netherlands <- ne_countries(scale = "medium", country = "Netherlands", returnclass = "sf")
df$longitude  = c(4.9041, 5.4778, 5.0803, 4.8681, 5.1214) 
df$latitude  = c(52.3676, 51.4416, 52.0306, 52.0282, 52.0907)
df_sf <- st_as_sf(df, coords = c("longitude", "latitude"), crs = 4326)
ggplot() +
  geom_sf(data = netherlands, fill = "gray", color = "black") +
  geom_sf(data = df_sf, aes(color = `2017 mean`), size = 5) +
  scale_color_gradientn(
    colors = c("green", "orange", "red"),
    name = "MDMA (mg/ 1000 people/day)"
  ) +
  geom_text(
    data = df,
    aes(x = longitude, y = latitude, label = city),
    vjust = ifelse(df$city == "Oudewater", 1.5, -1.5),
  ) +
  coord_sf(xlim = c(4, 7), ylim = c(51, 53)) +
  labs(title = "Mean levels of MDMA in wastewater in 2017 (mg/1000 people/day") +
  theme_minimal()
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
